Blum axioms and nondeterministic computation of functions

### 2015-12-15 paper.tex
paper.tex: Discuss acceptable Gödel numberings.

@@

@@

>>>>

\usepackage{amsmath}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

@@

\begin{definition}
    Let $\mathcal P$ be the the set of partial computable functions.
    A \emph{Gödel numbering}
    is a function $\phi: \{0, 1\}^* \to \mathcal P$
    that is surjective.
\end{definition}

A Gödel numbering formalizes the concept of associating a word $w$ of $\{0, 1\}^*$
with a partial recursive function $f$ of $\mathcal P$.
We can think of $w$ as being the ``source code'' for the ``program'' $f$;
we denote $f = \phi(w)$ by $\phi_w$.

Because Gödel numberings are infinite mathematical objects,
Turing machines cannot manipulate them directly.
We thus need some technique to manipulate the numbering indirectly.

Call $\pi$ the standard Gödel numbering
that corresponds to the standard representation of Turing machines;
that is, if $w$ is the encoding of a Turing machine $M$,
then $\pi_w$ is the partial function computed by $M$.
We can manipulate $\pi$ indirectly, for instance,
by means of a universal Turing machine $U$.
Given a program $w$ and an input $x$,
if $\pi_w(x)$ exists,
then the output of the machine $U$ when its input is the pair $(w, x)$
is exactly $\pi_w(x)$.
Therefore,
even tough we did not directly manipulate the numbering $\pi$
or calculate $\pi_w$,
we did manage to indirectly compute the value of $\pi_w(x)$.
This is the \emph{universal machine theorem}.

We can also manipulate the source code directly.
Suppose $w$ represents a Turing machine that computes a function of two variables $f$.
Given two inputs $x$ and $y$, we can compute $f(x, y) = \pi_w(x, y)$
using the universal Turing machine.
If we are given a single input $x$ and leave the second variable free,
we have a function $g$ of one variable, such that
\begin{equation*}
    g(y) = f(x, y).
\end{equation*}
That is, we fixed the first input to be $x$.

This function $g$ is computable;
in fact, we can obtain a source code $w'$ for $g$ from $w$
--- just modify the machine represented by $w$
so that, on input $y$,
it writes $x$ on the tape and runs the original algorithm.%
\footnote{
    Notice we are not manipulating the machine directly,
    only its representation as a string of bits.
}
This kind of modification can be done for any source code $w$
and any input $x$;
there is a computable function $\sigma$ of two variables
that is defined for every $w$ and $x$ such that,
for every $y$,
$\pi_w(x, y)$ is defined if and only if $\pi_{\sigma(w, x)}(y)$ is defined,
and, in this case,
\begin{equation*}
    \pi_w(x, y) = \pi_{\sigma(w, x)}(y).
\end{equation*}
This is the \emph{$S_{mn}$ theorem}.%
\footnote{
    The ``full version'' of this theorem asserts that,
    for any $m$ and $n$,
    there is a computable function $\sigma$ that,
    given a program $x$ for a function of $m + n$ variables
    and $n$ chosen values $y_1, \dots, y_m$,
    \begin{equation*}
        \pi_{\sigma(x, y_1, \dots, y_m)}(z_1, \dots, z_n) =
            \pi_x(y_1, \dots, y_m, z_1, \dots, z_n);
    \end{equation*}
    hence its name \cite[p.~23]{Rogers1987}.
}

These two theorems are all that Blum requires from a Gödel numbering
to define his axiomatic notion of complexity.
We have a special name (due to Rogers) for such numbering.

\begin{definition}
    An \emph{acceptable Gödel numbering}
    is a Gödel numbering $\phi$
    that satisfies both the universal machine theorem
    and the $S_{mn}$ theorem \cite[p.~41]{Rogers1987}.
    That is, there are computable functions $f$ and $\sigma$
    that satisfies
    \begin{itemize}
        \item For every $w$ and $x$,
            $f(w, x)$ is defined if and only if $\phi_w(x)$ is defined,
            and, in this case,
            \begin{equation*}
                f(w, x) = \phi_w(x);
            \end{equation*}
        \item $\sigma$ is defined for every $w$ and $x$,
            and, for every $y$,
            $\phi_w(x, y)$ is defined if and only if $\phi_{\sigma(w, x)}(y)$ is defined,
            and, in this case,
            \begin{equation*}
                \phi_w(x, y) = \phi_{\sigma(w, x)}(y).
            \end{equation*}
    \end{itemize}
\end{definition}

@@

\bibliographystyle{sbc/sbc}
\bibliography{bib/bibliography}
